---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# posologyr

<!-- badges: start -->
<!-- badges: end -->

## Overview

The goal of posologyr is to provide free tools (as in free speech) to individualize drug treatments by taking advantage of population pharmacokinetics, patient characteristics, and the results of therapeutic drug monitoring.

Posologyr offers functions for estimating pharmacokinetic parameters:

+ Estimation of the a priori distribution of the population pharmacokinetic parameters by Monte-Carlo simulations
+ Estimation of individual parameters Maximum A Posteriori (MAP), or Empirical Bayes Estimates (EBE), from the results of therapeutic drug monitoring
+ Estimation of the posterior distribution of individual pharmacokinetic parameters by Markov Chain Monte Carlo (MCMC)

Posologyr offers dosage optimisation functions based on estimated parameters:

+ Determination of the time needed to reach a target trough concentration (Cmin)
+ Determination of the optimal dose to reach a target AUC
+ Determination of the optimal dose to reach a target concentration at any given time

Posologyr requires a population pharmacokinetic model written in the specific language of RxODE, with an exponential model of inter-individual variability.

## Installation

You can install the development version of posologyr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("levenc/posologyr")
```
## Example

This example of Bayesian dosage adaptation is based on a fictitious population pharmacokinetic model of tobramycin, and data of therapeutic drug monitoring.

Patient data (here: Michel's results) should be organised in a dataframe that follows RxODE (or NONMEM) conventions.
```{r Michel}
library(posologyr)

df_michel <- data.frame(ID=1,TIME=c(0.0,0.5,1.0,14.0),DV=c(NA,NA,25.0,5.5),AMT=c(1000,-1000,0,0),EVID=c(10102,10102,0,0),DUR=c(0.5,0.5,NA,NA),CLCREAT=80,WT=65)
df_michel
```

Here the TIME is in hours. Following the RxODE conventions, the AMT is a rate in milligram/hour for the zero-order infusion of duration 0.5 hours. In this example 500 mg of tobramycin are administered over 30 minutes.

The tobramycin sample model is supplied with posologyr. It is a list of elements that includes all the necessary information, such as the structural model, the required covariates, and the variance covariance matrix.

Following the same structure, new models can be implemented.

```{r prior model}
# Fictional tobramycin 2cpt: example
mod_tobramycin_2cpt_fictional <- list(
  ppk_model   = RxODE::RxODE({
    centr(0) = 0
    ke = TVke*(CLCREAT/67.8)^0.89*(WT/66.4)^-1.09
    V  = TVV*(WT/66.4)^0.80
    Cc  = centr/V;
    d/dt(centr)  = - ke*centr - k12*centr + k21*periph;
    d/dt(periph) =            + k12*centr - k21*periph;
    d/dt(AUC)    =   Cc;
  }),
  error_model = error_model_comb1,
  pk_prior    = list( name = c('TVke','TVV','k12','k21'),
                      reference = c(TVke=0.21, TVV=19.8, k12=0.041, k21=0.12),
                      Omega = matrix(c(0.08075, 0      ,  0, 0,
                                       0      , 0.01203,  0, 0,
                                       0      , 0      ,  0, 0,
                                       0      , 0      ,  0, 0),
                                     ncol=4,byrow=TRUE)),
  covariates  = c("CLCREAT","WT"),
  xi          = c(additive_a = 0, proportional_b = 0.198))
```

The load_ppk_model() function initialises the objects to be used by posologyr. 

```{r load_ppk_model}
load_ppk_model(mod_tobramycin_2cpt_fictional,df_michel)
```

The typical values of the population parameters are available in the prior model

```{r TV parameters}
prior_ppk_model$pk_prior$reference
```

The MAP estimates of the individual parameters can be computed easily

```{r MAP individual parameters}
poso_estim_map()
```

One can then plot: 

```{r plots}
#The typical population profile,with the individual covariates, 
#over the individual concentration observations.
poso_plot_pop()
#The individual Maximum A Posteriori prediction
#over the individual observations
poso_plot_map()
#The individual posterior distribution over the individual observations
poso_plot_mcmc() 
```

The optimal dose can be estimated if, for example, a concentration of 30 mg/l is to be reached half an hour after the end of the infusion.

```{r optimal amount}
poso_dose_ctime(time_c = 1,duration = .5,target_ctime = 30)
```

To further optimize the dosage, it is now possible to determine the time needed to reach a Cmin < 0.5 mg/l, after an infusion of 620 mg over 30 minutes.

```{r optimal dosing}
poso_time_cmin(dose = 620, duration = 0.5, target_cmin = 0.5)
```

To conclude, the administration of 620 mg every 48 hours can be advisable for Michel.

## Acknowledgments
Posologyr's estimation functions were based on Marc Lavielle's code available at http://shiny.webpopix.org/mcmc/bayes1/

Posologyr relies heavily on the excellent RxODE package
