---
title: "A priori dose selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A priori dose selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

```{r chunkname, echo=-1}
data.table::setDTthreads(2)
```

```{r setup}
library(posologyr)
```

# Introduction
First dose selection for a critical care patient treated with amikacin for suspected ventilator-associated pneumonia. Population pharmacokinetic (ppk) model form Burdet et al. 2015.

```{r model}
mod_amikacin_Burdet2015 <- list(
  ppk_model   = rxode2::rxode({
    centr(0) = 0;
    TVCl  = THETA_Cl*(CLCREAT4H/82)^0.7;
    TVVc  = THETA_Vc*(TBW/78)^0.9*(PoverF/169)^0.4;
    TVVp  = THETA_Vp;
    TVQ   = THETA_Q;
    Cl    = TVCl*exp(ETA_Cl);
    Vc    = TVVc*exp(ETA_Vc);
    Vp    = TVVp*exp(ETA_Vp);
    Q     = TVQ *exp(ETA_Q);
    ke    = Cl/Vc;
    k12   = Q/Vc;
    k21   = Q/Vp;
    Cc    = centr/Vc;
    d/dt(centr)  = - ke*centr - k12*centr + k21*periph;
    d/dt(periph) =            + k12*centr - k21*periph;
    d/dt(AUC)    =   Cc;
  }),
  error_model = function(f,sigma){
    g <- sigma[1] + sigma[2]*f
    return(g)
  },
  theta = c(THETA_Cl=4.3, THETA_Vc=15.9, THETA_Vp=21.4,THETA_Q=12.1),
  omega = lotri::lotri({ETA_Cl + ETA_Vc + ETA_Vp + ETA_Q ~
      c(0.1,
        0.01     ,   0.05 ,
        0.01     ,   0.02 ,   0.2  ,
        -0.06    ,   0.004,   0.003,    0.08)}),
  covariates  = c("CLCREAT4H","TBW","PoverF"),
  sigma       = c(additive_a = 0.2, proportional_b = 0.1))
```

# A priori dose selection

## Patient record

Before the first administration, no concentration information is available. The patient record contains only the information required to fill in the covariates of the model: 

* CLCREAT4H: 4-h creatinine clearance in ml/min
* TBW: Total body weight in kg
* PoverF: PaO2/FIO2 ratio in mmHg

```{r prior_patientA}
df_patientA <- data.frame(ID=1,TIME=0,
                                DV=0,
                                EVID=0,
                                AMT=0,
                                CLCREAT4H=50,TBW=62,PoverF=169)
df_patientA
```

## Optimal dose selection

In the absence of measured concentrations, the optimal dose in mg to achieve a concentration of 80 mg/l one hour after the start of the 30-minute infusion is 
determined from the typical profile of the ppk model.

```{r prior_Cmax_optim_patientA}
prior_dose <- poso_dose_conc(dat=df_patientA,
                             prior_model=mod_amikacin_Burdet2015,
                             time_c = 1,                        #30 min after a  
                             duration = 0.5,                    #30 min infusion
                             target_conc = 80)
prior_dose
```

## Time required to reach the target Cmin

Following this dose, the time in hours required to reach a target Cmin concentration of 2.5 mg/l can be estimated.

```{r prior_Cmin_optim_patientA}
poso_time_cmin(dat=df_patientA,
               prior_model=mod_amikacin_Burdet2015,
               dose = prior_dose$dose,
               duration = 0.5,                                  #30 min infusion
               target_cmin = 2.5)
```

## Plotting the selected dosage

The selected dose can be simulated and plotted. By setting `n_simul = 0`, the `poso_simu_pop()` function produces a compiled `rxode2` model without inter-individual variability, using typical population parameter values and individual covariates from the patient record.

```{r prior_plot_model}
# generate a model using the individual covariates 
simu_patA      <- poso_simu_pop(dat=df_patientA,
                                prior_model=mod_amikacin_Burdet2015,
                                n_simul = 0)
```

Observations and a 30-minutes infusion of the optimal dose are added to the `rxode2` model by updating the [rxode2 event table](https://nlmixr2.github.io/rxode2/articles/rxode2-event-table.html).

```{r}
simu_patA$model$time <- seq(0,20,b=0.1)
simu_patA$model$add.dosing(dose=prior_dose$dose,rate=prior_dose$dose/0.5)
```

Plotting the simulated scenario.

```{r}
plot(simu_patA$model,Cc)
```

The resulting plot can be further augmented with `ggplot2`. For example, by adding an horizontal ribbon showing the 60-80 mg/l target interval of 1 h peak concentration, and a vertical dashed line marking 1 hour.

```{r prior_plot}
plot(simu_patA$model,Cc) + 
  ggplot2::ylab("Central concentration") +
  ggplot2::geom_vline(xintercept=1, linetype="dashed") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin=60, ymax=80),
                       fill="seagreen",show.legend = FALSE, alpha=0.15)
```

For a typical patient (i.e. with a PK profile typical of the model population), the selected dose meets the peak concentration target.
