---
title: "Case study: vancomycin dose adjustment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: vancomycin dose adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>"
)
```

```{r setup}
library(posologyr)
library(magrittr)
```

# Introduction
This case study presents the dosage adjustment for a patient treated with vancomycin for methicillin-resistant Staphylococcus aureus blood stream infection, using the population pharmacokinetic (ppk) model of Goti et al. `mod_vancomycin_2cpt_Goti2018`.

# Discontinuous intravenous infusion

## Patient record with TDM data

The dosage individualized using the results of therapeutic drug monitoring (TDM). See `vignette("patient_data_input")` for more details regarding the patient records.

```{r tdm_patientB}
df_patientB_tdm <- data.frame(ID=1,TIME=c(0.0,2,13.0,24.2,26.2,48),
                              DV=c(NA,NA,12,NA,NA,10.5),
                              AMT=c(1000,-1000,0,700,-700,0),
                              EVID=c(10101,10101,0,10101,10101,0),
                              CLCREAT=65,WT=70,DIAL=0)
df_patientB_tdm
```

## Apply the selected prior ppk model to patient data

The patient data and the ppk model are combined into a single object by `posologyr()`.

```{r posologyr_tdm_patientB}
patB_tdm_vanco <- posologyr(mod_vancomycin_2cpt_Goti2018,
                           df_patientB_tdm)
```

## Estimate the MAP individual parameters

```{r estim_map}
patB_map <- poso_estim_map(patB_tdm_vanco)
patB_map
```

## Plot the individual pharmacokinetic profile

The individual pharmacokinetic profile is plotted using the RxODE model provided by the `poso_estim_map()` function.

```{r map_plot_tdm}
#Get the observations from the patient record
indiv_obs             <- patB_tdm_vanco$tdm_data[,c("DV","TIME")]
names(indiv_obs)      <- c("value","time")

#MAP profile
map_model_patB        <- patB_map$model
map_model_patB$time   <- seq(0,48,by=0.1)

#Overlay the MAP profile and the observations
plot(map_model_patB,Cc) +
  ggplot2::ylab("Central concentration") +
  ggplot2::geom_point(data=indiv_obs, size= 3, na.rm=TRUE)

```

## Get the AUC24 from the MAP model

Considering a MIC of 1 mg/L, the target AUC over 24 hours (AUC24) is 400 mg.h/L.

```{r AUC_map_dose}
#AUC 0_24
AUC_map_first_dose <- map_model_patB$AUC[which(map_model_patB$time == 24)]
AUC_map_first_dose

#AUC 24_48
AUC_map_second_dose <- map_model_patB$AUC[which(map_model_patB$time == 48)] - AUC_map_first_dose
AUC_map_second_dose
```

The current dosage does not meet the target AUC.

## Optimal maintenance dose selection a posteriori

The maintenance dose needed to achieve an AUC24 of 400 mg.h/L is estimated by simulating a multiple dose regimen over enough administrations (11 consecutive administrations, with `add_dose=10`) to approximate the steady-state.

```{r optim_maintenance_dose}
maintenance_dose_map <- poso_dose_auc(patB_tdm_vanco,
                                      time_auc=24,
                                      starting_time=24*9,
                                      interdose_interval=24,
                                      add_dose=10,
                                      duration=2,
                                      target_auc=400)
maintenance_dose_map
```

## Optimal loading dose selection a posteriori

The ideal loading dose achieves the target steady-state concentration on the first administration. By simulating the administration of the maintenance dose to steady state, the Cmin at steady state can be determined.

```{r cmin_ss}
#event table: multiple dose regimen until steady state
et_ss <- RxODE::et(amt=maintenance_dose_map,dur=2,ii=24,until=240)
et_ss %<>% RxODE::et(seq(0,240,by=0.1))

#Simulate with the event table and the individual MAP parameters
map_optim_model_patB <- RxODE::rxSolve(object=patB_tdm_vanco$ppk_model,
                                      params=patB_map$model$params,
                                      events=et_ss)

#Cmin at steady-state
target_cmin_ss <- map_optim_model_patB$Cc[which(map_optim_model_patB$time == 24*9)]
target_cmin_ss
```

The dose needed to achieve the steady-state Cmin following the first administration can now be identified.

```{r loading_dose}
loading_dose_map <- poso_dose_ctime(patB_tdm_vanco,
                                    time_c=24,
                                    duration=2,
                                    target_conc=target_cmin_ss)
loading_dose_map
```

## Simulate and plot the optimal a posteriori dosage

```{r plot_optim_dosage}
#event table: loading dose followed by maintenance dose once daily
et_load         <- RxODE::et(amt=loading_dose_map,dur=2)
et_maint        <- RxODE::et(amt=maintenance_dose_map,dur=2,ii=24,until=240)
et_optim_dosage <- seq(et_load,et_maint) %>% RxODE::et(seq(0,240,by=0.1))

#Simulate with the event table and the individual MAP parameters
map_optim_model_patB <- RxODE::rxSolve(object=patB_tdm_vanco$ppk_model,
                                      params=patB_map$model$params,
                                      events=et_optim_dosage)

#Plot the results of the simulation
plot(map_optim_model_patB,Cc) + 
  ggplot2::ylab("Central concentration")
```

## Get the AUC24 for the optimal a posteriori dosage

```{r AUC_optim_map_cmin_ss}
#AUC 0_24
AUC_loading_dose <- map_optim_model_patB$AUC[which(map_optim_model_patB$time == 24)]
AUC_loading_dose

#AUC 24_48
AUC_maintenance_dose <- map_optim_model_patB$AUC[which(map_optim_model_patB$time == 48)] -
  AUC_loading_dose
AUC_maintenance_dose
```

The target AUC24 are now met.

# Continuous intravenous infusion

The maintenance dose for a continuous intravenous infusion can be determined by setting the `duration` of the infusion equal to the `interdose_interval`.

```{r continuous_infusion}
#Target Css for a continuous infusion
target_css <- 400/24

maintenance_dose_map_ci <- poso_dose_ctime(patB_tdm_vanco,
                                           time_c=24*9.9,
                                           interdose_interval=24,
                                           add_dose=10,
                                           duration=24,
                                           target_conc=target_css)
maintenance_dose_map_ci
```

## Simulate and plot the optimal a posteriori dosage

```{r ci_plots}
#event table: loading dose followed by continuous infusion
et_maint_ci <- RxODE::et(time=2,
                         amt=maintenance_dose_map_ci,
                         dur=24,
                         ii=24,
                         until=240)
et_ci       <- et_maint_ci %>% RxODE::add.dosing(dose=loading_dose_map,dur=2)
et_ci       %<>% RxODE::et(seq(0,240,by=0.1))

#Simulate with the event table and the individual MAP parameters
map_optim_model_patB_ci  <- RxODE::rxSolve(object=patB_tdm_vanco$ppk_model,
                                          params=patB_map$model$params,
                                          events=et_ci)

#Plot the results of the simulation
plot(map_optim_model_patB_ci,Cc) + 
  ggplot2::ylab("Central concentration")
```
