---
title: "Case study: amikacin dose adjustment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: amikacin dose adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(posologyr)
```

# Introduction
This case study presents the dosage adjustment for a critical care patient treated with amikacin for suspected ventilator-associated pneumonia, using the population pharmacokinetic (ppk) model of Burdet et al. `mod_amikacin_2cpt_Burdet2015`.

# A priori drug dosing

## Patient record

Before the first administration, no concentration information is available. The patient record contains only the information required to fill in the covariates of the model, see `help("mod_amikacin_2cpt_Burdet2015")`.

```{r prior_patientA}
df_patientA_prior <- data.frame(ID=1,TIME=0,
                                DV=0,
                                EVID=0,
                                AMT=0,
                                CLCREAT4H=50,TBW=62,PoverF=169)
df_patientA_prior
```

## Apply the selected prior ppk model to patient data

The `posologyr()` function combines the patient record and the ppk model into a single object.

```{r posologyr_prior_patientA}
patA_prior_amik <- posologyr(mod_amikacin_2cpt_Burdet2015,
                             df_patientA_prior)
```

## Optimal dose selection

In the absence of measured concentrations, the optimal dose in mg to achieve a so-called peak concentration of 80 mg/l (one hour after the start of the 30-minute infusion) is determined from the typical profile of the ppk model.

```{r prior_Cmax_optim_patientA}
prior_dose <- poso_dose_ctime(patA_prior_amik,
                              time_c = 1,
                              duration = 0.5,
                              target_conc = 80)
prior_dose
```

## Time required to reach the target Cmin

With this dose, the time in hours required to reach a target Cmin concentration of 2.5 mg/l can be estimated.

```{r prior_Cmin_optim_patientA}
prior_timecmin <- poso_time_cmin(patA_prior_amik,
                                 dose = prior_dose,
                                 duration = 0.5,
                                 target_cmin = 2.5)
prior_timecmin
```

## Simulate and plot the a priori dosage

An event table is created to simulate the selected dosage with RxODE.

```{r prior_plot_et}
# infusion of prior_dose mg over 0.5 hours at time 0
event_table_priordose <- RxODE::et(time=0,amt=prior_dose,dur=0.5)
event_table_priordose$add.sampling(seq(0,24,by=0.2))
```

By setting `n_simul = 0`, the `poso_simu_pop()` function produces an RxODE model that incorporates typical population parameter values and individual covariates from the patient record.

```{r prior_plot_model}
prior_param_patA      <- poso_simu_pop(patA_prior_amik,n_simul = 0)$model$params

prior_dose_model_patA <- RxODE::rxSolve(object=patA_prior_amik$ppk_model,
                                  params=prior_param_patA,
                                  events=event_table_priordose)
```

The resulting model is plotted with an horizontal ribbon showing the 60-80 mg/l target interval of 1 h peak concentration, and a vertical dashed line marking 1 hour.

```{r prior_plot}
plot(prior_dose_model_patA,Cc) + 
  ggplot2::ylab("Central concentration") +
  ggplot2::geom_vline(xintercept=1, linetype="dashed") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin=60, ymax=80),
                       fill="seagreen",show.legend = FALSE, alpha=0.15)
```

For a patient with a pharmacokinetic profile typical of the model population, the selected dose meets the peak concentration target.

# A posteriori drug dosing

## Patient record with TDM data

After the first administration, the dosage selection can be refined using the results of therapeutic drug monitoring (TDM). The new data are added to the patient file, see `vignette("patient_data_input")` for more details.

```{r tdm_patientA}
df_patientA_tdm <- data.frame(ID=1,TIME=c(0,0.5,1,6),
                              DV=c(NA,NA,58,14),
                              EVID=c(10101,10101,0,0),
                              AMT=c(4160,-4160,0,0),
                              CLCREAT4H=50,TBW=62,PoverF=169)
df_patientA_tdm
```

## Apply the selected prior ppk model to patient data

The updated patient record and the ppk model are combined into a single object by `posologyr()`.

```{r posologyr_tdm_patientA}
patA_tdm_amik <- posologyr(mod_amikacin_2cpt_Burdet2015,
                           df_patientA_tdm)
```

## Plot the observations from TDM over the simulated a priori dosage

The observation from TDM are taken from the patient record, and renamed to match the plot of the prior dose model.

```{r prior_plot_tdm}
indiv_obs           <- patA_tdm_amik$tdm_data[,c("DV","TIME")]
names(indiv_obs)    <- c("value","time")

plot(prior_dose_model_patA,Cc) + 
  ggplot2::ylab("Central concentration") +
  ggplot2::geom_point(data=indiv_obs, size= 3, na.rm=TRUE) +
  ggplot2::geom_vline(xintercept=1, linetype="dashed") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin=60, ymax=80),
                       fill="red4",show.legend = FALSE, alpha=0.15)
```

The observations are not close to the a priori population profile. The target peak concentration was not met.

## Estimate the MAP individual parameters

The maximum a posteriori (MAP) individual parameters are estimated.

```{r estim_map}
patA_map <- poso_estim_map(patA_tdm_amik)
patA_map
```

## Plot the individual pharmacokinetic profile

The individual pharmacokinetic profile is plotted using the RxODE model provided by the `poso_estim_map()` function.

```{r map_plot_tdm}
map_model_patA      <- patA_map$model
map_model_patA$time <- seq(0,24,by=0.2) 

plot(map_model_patA,Cc) + 
  ggplot2::ylab("Central concentration") +
  ggplot2::geom_point(data=indiv_obs, size= 3, na.rm=TRUE) +
  ggplot2::geom_vline(xintercept=1, linetype="dashed") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin=60, ymax=80),
                       fill="red4",show.legend = FALSE, alpha=0.15)
```

The MAP profile matches the observations.

## Time required to reach the target Cmin following the first administration

With the MAP estimates of the individual parameters, the prediction of the time in hours needed before reaching the target Cmin can be updated.

```{r map_Cmin_priordose_patientA}
map_timecmin_priordose <- poso_time_cmin(patA_tdm_amik,
                                         dose = prior_dose,
                                         duration = 0.5,
                                         target_cmin = 2.5)
map_timecmin_priordose
```

## Optimal dose selection a posteriori

An optimal a posteriori dose to achieve a peak concentration of 80 mg/l is determined using the MAP estimates.

```{r map_Cmax_optim_patientA}
map_dose <- poso_dose_ctime(patA_tdm_amik,
                            time_c = 1,
                            duration = 0.5,
                            target_conc = 80)
map_dose
```

## Interdose interval selection a posteriori

The time required to reach a Cmin of 2 mg/L after a single administration is determined using the MAP estimates.

```{r map_Cmin_optim_patientA}
map_timecmin <- poso_time_cmin(patA_tdm_amik,
                               dose = map_dose,
                               duration = 0.5,
                               target_cmin = 2)
map_timecmin
```

## Simulate and plot the optimal a posteriori dosage

### Single administration

```{r map_dose_plot}
event_table_mapdose <- RxODE::et(time=0,amt=map_dose,dur=0.5)
event_table_mapdose$add.sampling(seq(0,24,by=0.1))

map_param_patA      <- patA_map$model$params

map_dose_model_patA <- RxODE::rxSolve(object=patA_tdm_amik$ppk_model,
                                  params=map_param_patA,
                                  events=event_table_mapdose)

plot(map_dose_model_patA,Cc) + 
  ggplot2::ylab("Central concentration") +
  ggplot2::geom_vline(xintercept=1, linetype="dashed") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin=60, ymax=80),
                       fill="seagreen",show.legend=FALSE, alpha=0.15)
```

### Multiple administrations

```{r map_dose_ss_plot}
event_table_mapdose_ss <- RxODE::et(amt=map_dose,dur=0.5,ii=map_timecmin,addl=2)
event_table_mapdose_ss$add.sampling(seq(0,3.5*map_timecmin,by=0.1))

map_dose_model_patA_ss <- RxODE::rxSolve(object=patA_tdm_amik$ppk_model,
                                  params=map_param_patA,
                                  events=event_table_mapdose_ss)

plot(map_dose_model_patA_ss,Cc) + 
  ggplot2::ylab("Central concentration") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin=60, ymax=80),
                       fill="seagreen",show.legend = FALSE, alpha=0.15) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin=0, ymax=2.5),
                       fill="seagreen",show.legend = FALSE, alpha=0.15)
```
