---
title: "Patient data input"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Patient data input}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
This describes the structure of patient records usable by `posologyr`.

# Structure
Data for input into `posologyr` is the same type of data input for rxode2. As stated in the [rxode2 documentation (rxode2 datasets)](https://nlmixr2.github.io/rxode2/articles/rxode2-datasets.html), it is also similar to [data for NONMEM](https://doi.org/10.1002/psp4.12404).

The patient dataset is a table of sequential event records, in which each line is an event. A description of the different event types is available in the [rxode2 documentation (rxode2 Event Types)](https://nlmixr2.github.io/rxode2/articles/rxode2-event-types.html).

Below is a minimal working example:

```{r minimal_patient_df}
data.frame(ID=1,
           TIME=c(0.0,3),
           DV=c(NA,60.0),
           AMT=c(1000,0),
           EVID=c(101,0))
```

## Required fields and data

TIME
: Dosing times, and sampling times of therapeutic drug monitoring (TDM). The units depend on the specification of the population pharmacokinetics (ppk) model.

AMT
: Amount of drug administered. The units depend on the specification of the ppk model.

EVID
: Event type. Must be 0 for all observations (concentrations from TDM).

DV 
: For concentrations. Must be NA when EVID is not 0.

Covariates
: Every covariate defined in the prior `posologyr` model. The column labels must match the **covariate** vector from the ppk model.

OCC
: Occasions. Required for models with inter-occasion variability.

# Common use cases
## A priori dose adjustment
Before dosing: the simplest patient record is a single line dataframe, with individual patient covariates, and all other columns (TIME, DV, AMT, EVID) set to zero.

```{r before_dosing_df}
data.frame(ID=1,
           TIME=0,
           DV=0,
           AMT=0,
           EVID=0,
           COVAR1=c("X"),
           COVAR2=c("Y"))
```

## Oral administration or IV bolus
`EVID = 1` for an instantaneous administration in the first compartment defined in the ppk model: either the central compartment for an IV bolus, or a depot compartment for an oral administration.

```{r oral_IV_bolus_df}
data.frame(ID=1,
           TIME=c(0.0,3),
           DV=c(NA,60.0),
           AMT=c(1000,0),
           EVID=c(1,0),
           COVAR1=c("X"),
           COVAR2=c("Y"))
```

## Intermittent infusion
`EVID = 1` for a bolus infusion, administered in the first compartment defined in the ppk model.

`DUR` defines its duration.

`AMT` is the amount administered over the duration `DUR`.

```{r intermittent_infusion_df}
data.frame(ID=1,
           TIME=c(0.0,1.0,14.0),
           DV=c(NA,25.0,5.5),
           AMT=c(1000,0,0),
           DUR=c(0.5,NA,NA),
           EVID=c(1,0,0),
           COVAR1=c("X"),
           COVAR2=c("Y"))
```

## Inter-occasion variability
For a given occasion, the value of `OCC` must be repeated for each row. `OCC` must be specified for all the rows in the table.

```{r iov_df}
data.frame(ID=1,
           TIME=c(0.0,1.0,14.0,24.0,25.0,36.0),
           DV=c(NA,25.0,5.5,NA,30.0,6.0),
           AMT=c(1000,0,0,1000,0,0),
           DUR=c(0.5,NA,NA,0.5,NA,NA),
           EVID=c(1,0,0,1,0,0),
           OCC=c(1,1,1,2,2,2),
           COVAR1=c("X"),
           COVAR2=c("Y"))
```
