---
title: "User defined models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User defined models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Introduction
This describes the structure of prior models usable by `posologyr` and illustrates how to define new models from published population pharmacokinetic (popPK) models.

# Structure
A `posologyr` prior popPK model is a list of six objects:

ppk_model 
: A RxODE model implementing the structural population pharmacokinetics model with the individual model (i.e. the model of inter-individual variability) and the covariates

error_model 
: A function of the residual error model

theta
: A named vector of the population estimates of the fixed effects parameters (called THETAs, following [NONMEM terminology](https://doi.org/10.1002/psp4.12404))

omega 
: A named square variance-covariance matrix of the population parameters inter-individual variability

covariates 
: A character vector of the covariates of the model

xi 
: The estimates of the parameters of the residual error model

# Definition of a prior model through an example
The model to implement is a two-compartment popPK model of vancomycin derived from a retrospective study with a cohort of over 1,800 patients ([doi:10.1097/FTD.0000000000000490](https://doi.org/10.1097/FTD.0000000000000490)).

## ppk_model
A model defined in the RxODE::RxODE mini-language. `posologyr` needs a structural model, defined with either differential or algebraic equations, and an invidivual model.

The concentration in the central compartment *must* be named Cc.

The differential function `d/dt(AUC) =  Cc;` is needed for the optimisation function `poso_dose_auc()`.

```{r ppk_model}
ppk_model   = RxODE::RxODE({
    centr(0) = 0;
    TVCl  = THETA_Cl*(CLCREAT/120)^0.8*(0.7^DIAL);
    TVVc  = THETA_Vc*(WT/70)          *(0.5^DIAL);
    TVVp  = THETA_Vp;
    TVQ   = THETA_Q;
    Cl    = TVCl*exp(ETA_Cl);
    Vc    = TVVc*exp(ETA_Vc);
    Vp    = TVVp*exp(ETA_Vp);
    Q     = TVQ;
    ke    = Cl/Vc;
    k12   = Q/Vc;
    k21   = Q/Vp;
    Cc    = centr/Vc;
    d/dt(centr)  = - ke*centr - k12*centr + k21*periph;
    d/dt(periph) =            + k12*centr - k21*periph;
    d/dt(AUC)    =   Cc;
  })
```

## error_model
A function of the residual error model, taking two arguments: the simulated concentrations, and a vector **xi** of the estimates of the parameters for the residual error model.

```{r error_model}
error_model <- function(f,xi){     #additive model if xi[2] == 0
  g <- xi[1] + xi[2]*f             #proportional model if xi[1] == 0
  return(g)
}
```

## theta
The estimations of the parameters for the fixed effects of the model (THETA), in a named vector. The names must match the names used in **ppk_model**.

```{r theta}
theta = c(THETA_Cl=4.5, THETA_Vc=58.4, THETA_Vp=38.4, THETA_Q=6.5)
```

## omega
The variance-covariance matrix of the random effects (ETA) for the individual model. A symmetric matrix. The names must match the names used in **ppk_model**. The easiest way to define it is with `lotri::lotri`.

The estimates of the variances of the random effects can be given under different parameterizations depending on the authors.

* Standard deviation (SD): the square root of the variance, as returned by Monolix
* Coefficient of variation (CV): calculated as `sqrt(exp(SD^2)-1)`, the standard deviation can be computed back with `sqrt(log((CV^2)+1))`
* Full covariance matrix: the easiest to reuse, but rarely seen in articles

In the case of the [vancomycin model](https://doi.org/10.1097/FTD.0000000000000490), the estimates of between subject variability (BSV) are given as CV%. They must be converted to variances prior to their inclusion in **omega**.

|Parameter |CV%  (from the article)|SD       |Variance = SD^2|
|----------|:---------------------:|:-------:|:-------------:|
|BSV on CL |39.8                   | 0.383   | 0.147         |
|BSV on Vc |81.6                   | 0.714   | 0.510         |
|BSV on Vp |57.1                   | 0.531   | 0.282         |

```{r omega}
omega = lotri::lotri({ETA_Cl + ETA_Vc + ETA_Vp + ETA_Q ~
                          c(0.147,
                           0     ,  0.510 ,
                           0     ,       0,   0.282,
                           0     ,       0,       0,    0)})

```

The estimates of covariance (off diagonal) are sometimes given as coefficients of correlation between ETAs. The covariance between ETA_a and ETA_b can be computed with the following product: `standard_deviation(ETA_a) * standard_deviation(ETA_b) * correlation(ETA_a and ETA_b)`.

In this example, all covariances are egal to zero.

## covariates
The names of every covariate defined in **ppk_model**, in a character vector.

```{r covariates}
covariates  = c("CLCREAT","WT","DIAL")
```

## xi
The estimates of the parameters for the residual error model, in a vector.

```{r xi}
xi          = c(additive_a = 3.4, proportional_b = 0.227)
```

## Full model
The `posologyr` model is the list of all these objects.

```{r vancomycin_pososlogyr_model_list}
mymod_vancomyin_Goti2018 <- list(
  ppk_model   = RxODE::RxODE({
     centr(0) = 0;
    TVCl  = THETA_Cl*(CLCREAT/120)^0.8*(0.7^DIAL);
    TVVc  = THETA_Vc*(WT/70)          *(0.5^DIAL);
    TVVp  = THETA_Vp;
    TVQ   = THETA_Q;
    Cl    = TVCl*exp(ETA_Cl);
    Vc    = TVVc*exp(ETA_Vc);
    Vp    = TVVp*exp(ETA_Vp);
    Q     = TVQ;
    ke    = Cl/Vc;
    k12   = Q/Vc;
    k21   = Q/Vp;
    Cc    = centr/Vc;
    d/dt(centr)  = - ke*centr - k12*centr + k21*periph;
    d/dt(periph) =            + k12*centr - k21*periph;
    d/dt(AUC)    =   Cc;
  }),
  error_model = function(f,xi){
    g <- xi[1] + xi[2]*f
    return(g)
  },
  theta = c(THETA_Cl=4.5, THETA_Vc=58.4, THETA_Vp=38.4,THETA_Q=6.5),
  omega = lotri::lotri({ETA_Cl + ETA_Vc + ETA_Vp + ETA_Q ~
                          c(0.147,
                           0     ,  0.510 ,
                           0     ,       0,   0.282,
                           0     ,       0,       0,    0)}),
  covariates  = c("CLCREAT","WT","DIAL"),
  xi          = c(additive_a = 3.4, proportional_b = 0.227))
```

### Resulting R object

```{r vancomycin_pososlogyr_model}
mymod_vancomyin_Goti2018
```
