---
title: "Uncertainty of estimates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty of estimates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(posologyr)
```

# Introduction

The case study of amikacin dose adjustment `vignette("case_study_amikacin")` focuses on point estimates: the typical values of population parameters, and MAP estimates. From the same data, the uncertainty associated with the estimation of the parameters can be assessed.

# Population parameters

## Patient record and prior model from the amikacin case study

```{r prior_patientA}
df_patientA_prior <- data.frame(ID=1,TIME=0,
                                DV=0,
                                EVID=0,
                                AMT=0,
                                CLCREAT4H=50,TBW=62,PoverF=169)
df_patientA_prior
```

```{r posologyr_prior_patientA}
patA_prior_amik <- posologyr(mod_amikacin_2cpt_Burdet2015,df_patientA_prior)
```

## Distribution of the population parameters

The inter-individual variability of the pharmacokinetic parameters can be explored by simulation, using the `poso_simu_pop()` function to generate sets of ETAs. The resulting distribution of ETAs is used to simulate and plot the pk profiles.

```{r priordist_plot}
event_table_priordose <- RxODE::et(time=0,amt=2080,dur=0.5)
event_table_priordose$add.sampling(seq(0,24,by=0.2))

priordist_param_patA      <- poso_simu_pop(patA_prior_amik)$model$params

priordist_dose_model_patA <- RxODE::rxSolve(object=patA_prior_amik$ppk_model,
                                  params=priordist_param_patA,
                                  event_table_priordose)

plot(confint(priordist_dose_model_patA,"Cc", level=0.95),ylab="Central concentration")
```

# MAP estimation of the parameters

## Patient record with TDM and prior model from the amikacin case study

```{r tdm_patientA}
df_patientA_tdm <- data.frame(ID=1,TIME=c(0,0.5,1,6),
                              DV=c(NA,NA,60,14),
                              EVID=c(10101,10101,0,0),
                              AMT=c(4160,-4160,0,0),
                              CLCREAT4H=50,TBW=62,PoverF=169)
df_patientA_tdm
```

```{r posologyr_tdm_patientA}
patA_tdm_amik <- posologyr(mod_amikacin_2cpt_Burdet2015,df_patientA_tdm)
```

## Relative standard error (RSE) of the MAP estimates

The precision of the the maximum a posteriori (MAP) estimates can be computed by setting `return_rse = TRUE` when using the `poso_estim_map()` function; `$fim` is the Fisher Information Matrix. 

```{r estim_map}
patA_map <- poso_estim_map(patA_tdm_amik,return_fim = TRUE,return_rse = TRUE)
patA_map
```

RSE below 30% are usually considered acceptable.

# Conditionnal distribution of the individual parameters

## MCMC estimation
Beyond point estimates, the a posteriori distribution of the individual parameters can be computed by Markov Chain Monte Carlo (MCMC) estimation. The `poso_estim_mcmc()` function can be slow, the number of iterations of the Metropolis-Hastings algorithm can be tuned.

```{r estim_mcmc}
patA_mcmc <- poso_estim_mcmc(patA_tdm_amik)
```

## Plot of the distribution of the individual profiles

The distribution of the individual profiles obtained can be plotted. The dispersion is significantly lower than that observed for the a priori population profile.

```{r mcmc_plot_tdm}
mcmc_model_patA      <- patA_mcmc$model
mcmc_model_patA$time <- seq(0,24,by=0.2) 

plot(confint(mcmc_model_patA,"Cc", level=0.95),ylab="Central concentration")
```

## Plot of the iterations of the MCMC algorithm for each parameter

```{r fuzzy_caterpillars}
plot(patA_mcmc$eta[,1],type="l",ylab="ETA_Cl")
plot(patA_mcmc$eta[,2],type="l",ylab="ETA_Vc")
plot(patA_mcmc$eta[,3],type="l",ylab="ETA_Vp")
plot(patA_mcmc$eta[,4],type="l",ylab="ETA_Q")
```

## Variance-covariance matrix

The a priori variance-covariance matrix is the omega matrix of the population pharmacokinetic (ppk) model. The variance covariance matrix of the conditional distribution of ETAs is computed from the output of `poso_estim_mcmc()`.

```{r variance_covariance}
patA_tdm_amik$omega
var(patA_mcmc$eta)
```

## Prior and posterior distribution of the parameters

The density plots show the differences between the prior and the posterior distributions.

```{r density_plots}
prior_estimates <- poso_simu_pop(patA_tdm_amik)

plot(density(patA_mcmc$model$Cl),ylim = c(0, 0.5),main="Distributions of Cl")
lines(density(prior_estimates$model$Cl),col=2)
legend("topright",
       c("prior distribution", "posterior distribution"),
       col = 1:2,
       lty = 1)
plot(density(patA_mcmc$model$Vc),ylim = c(0, 0.2),main="Distributions of Vc")
lines(density(prior_estimates$model$Vc),col=2)
legend("topright",
       c("prior distribution", "posterior distribution"),
       col = 1:2,
       lty = 1)
plot(density(patA_mcmc$model$Vp),ylim = c(0, 0.06),main="Distributions of Vp")
lines(density(prior_estimates$model$Vp),col=2)
legend("topright",
       c("prior distribution", "posterior distribution"),
       col = 1:2,
       lty = 1)
plot(density(patA_mcmc$model$Q),ylim = c(0, 0.2),main="Distributions of Q")
lines(density(prior_estimates$model$Q),col=2)
legend("topright",
       c("prior distribution", "posterior distribution"),
       col = 1:2,
       lty = 1)
```
