---
title: "A posteriori dose selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A posteriori dose selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
library(rxode2)
setRxThreads(2L)  # limit the number of threads
```

```{r setup}
library(posologyr)
```

# Introduction
Dosage individualization for a critical care patient treated with amikacin for suspected ventilator-associated pneumonia, using the population pharmacokinetic (ppk) model of Burdet et al. 2015, using the data from therapeutic drug monitoring (TDM).

```{r model}
mod_amikacin_Burdet2015 <- function() {
    ini({
      THETA_Cl=4.3
      THETA_Vc=15.9
      THETA_Vp=21.4
      THETA_Q=12.1
      ETA_Cl + ETA_Vc + ETA_Vp + ETA_Q ~
        c(0.1,
          0.01     ,   0.05 ,
          0.01     ,   0.02 ,   0.2  ,
          -0.06    ,   0.004,   0.003,    0.08)
      add_sd <- 0.2
      prop_sd <- 0.1
    })
    model({
      TVCl  = THETA_Cl*(CLCREAT4H/82)^0.7
      TVVc  = THETA_Vc*(TBW/78)^0.9*(PoverF/169)^0.4
      TVVp  = THETA_Vp
      TVQ   = THETA_Q
      Cl    = TVCl*exp(ETA_Cl)
      Vc    = TVVc*exp(ETA_Vc)
      Vp    = TVVp*exp(ETA_Vp)
      Q     = TVQ *exp(ETA_Q)
      ke    = Cl/Vc
      k12   = Q/Vc
      k21   = Q/Vp
      Cp    = centr/Vc
      d/dt(centr)  = - ke*centr - k12*centr + k21*periph
      d/dt(periph) =            + k12*centr - k21*periph

      Cp ~ add(add_sd) + prop(prop_sd) + combined1()
    })
}

mod_amikacin_Burdet2015 <- mod_amikacin_Burdet2015()
```

# A posteriori dose selection

## Patient record with TDM data

After the first administration, the dosage selection can be refined using the results of TDM. See `vignette("patient_data_input")` for more details regarding the patient record.

```{r tdm_patientA}
df_patientA <- data.frame(ID=1,TIME=c(0,1,6),
                              DV=c(NA,58,14),
                              EVID=c(1,0,0),
                              AMT=c(2000,0,0),
                              DUR=c(0.5,NA,NA),
                              CLCREAT4H=50,TBW=62,PoverF=169)
df_patientA
```
The concentration measured 30 min after a 30 min infusion do not meet the target for a peak concentration; it is < 60 mg/L.

## Estimate the MAP individual parameters

The maximum a posteriori (MAP) individual parameters are estimated.

```{r estim_map}
patA_map <- poso_estim_map(dat=df_patientA,
                           prior_model=mod_amikacin_Burdet2015)
```

## Plot the individual pharmacokinetic profile

The individual pharmacokinetic profile can be plotted using the `rxode2` model provided by the `poso_estim_map()` function.

```{r map_plot_tdm}
#| fig.alt: >
#|   Plot of the individual PK profile
plot(patA_map$model,Cc)
```

## Time required to reach the target Cmin following the first administration

With the MAP estimates of the individual parameters, the prediction of the time needed before reaching the target Cmin can be updated.

```{r map_Cmin_priordose_patientA}
poso_time_cmin(dat=df_patientA,
               prior_model=mod_amikacin_Burdet2015,
               tdm = TRUE,
               target_cmin = 2.5)
```
The next dose (if needed) can be administered 33.9 hours following the first infusion.

## Optimal dose selection a posteriori

The optimal dose to achieve a peak concentration of 80 mg/l can be determined using the MAP estimates.

```{r map_Cmax_optim_patientA}
map_dose <- poso_dose_conc(dat=df_patientA,
                           prior_model=mod_amikacin_Burdet2015,
                           tdm=TRUE,
                           time_c = 35,               #target concentration at t = 35 h
                           time_dose = 34,            #dosing at t = 34 h
                           duration = 0.5,
                           target_conc = 80)
map_dose
```
The next dose should be 2450 mg.

## Interdose interval selection a posteriori

The optimal inter-dose interval to reach a Cmin of 2.5 mg/L before each dosing can be determined using the MAP estimates.

```{r map_Cmin_optim_patientA}
map_interval <- poso_inter_cmin(dat=df_patientA,
                                prior_model=mod_amikacin_Burdet2015,
                                dose = map_dose$dose,
                                duration = 0.5,
                                target_cmin = 2.5)
map_interval
```
The interval between doses should not be less than 38.6 hours to allow adequate elimination of amikacin between each infusion.
