---
title: "Classic posologyr models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classic posologyr models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Introduction
Originally, `posologyr` models were R lists; `posologyr` still uses this format internally, but it's often more useful to use the `rxode2` syntax, or even to import a NONMEM model using the `nonmem2rx` package.

This article describes the structure of classical `posologyr` models. It illustrates how to define them from published population models.

# Structure
A `posologyr` model is a named R list of the following items:

ppk_model 
: A `rxode2` model implementing the structural population model with the individual model (i.e. the model of inter-individual variability) and the covariates

error_model 
: A function of the residual error model, alternatively a named list of functions for multiple endpoints model `vignette("multiple_endpoints")`

theta
: A named vector of the population estimates of the fixed effects parameters (called THETAs, following [NONMEM terminology](https://doi.org/10.1002/psp4.12404))

omega 
: A named square variance-covariance matrix of the population parameters inter-individual variability

sigma 
: The estimates of the parameters of the residual error model

pi_matrix
: Optional. A named square variance-covariance matrix of the population parameters inter-occasion variability

covariates 
: A character vector of the covariates of the model

# Definition of a prior model through an example
The model in this example is a two-compartment ppk model of vancomycin derived from a retrospective study with a cohort of over 1,800 patients ([doi:10.1097/FTD.0000000000000490](https://doi.org/10.1097/FTD.0000000000000490)).

## ppk_model
A model defined in the `rxode2` mini-language. `posologyr` needs a structural model, defined with either differential or algebraic equations, and an individual model. Depending on model type, naming conventions are more or less strict:

**Single endpoint model** (e.g. most pharmacokinetic models)

 * The concentration in the central compartment *must* be named Cc.  

**Multiple endpoints model** (eg. PK-PD, parent-metabolite, blood-urine…)

 * The names of the endpoints are flexible, but they must be consistent with the names of the error models and their parameters.

The differential function `d/dt(AUC) =  Cc;` is needed for the optimization function `poso_dose_auc()`.

```{r ppk_model}
ppk_model   = rxode2::rxode({
    centr(0) = 0;
    TVCl  = THETA_Cl*(CLCREAT/120)^0.8*(0.7^DIAL);
    TVVc  = THETA_Vc*(WT/70)          *(0.5^DIAL);
    TVVp  = THETA_Vp;
    TVQ   = THETA_Q;
    Cl    = TVCl*exp(ETA_Cl);
    Vc    = TVVc*exp(ETA_Vc);
    Vp    = TVVp*exp(ETA_Vp);
    Q     = TVQ;
    ke    = Cl/Vc;
    k12   = Q/Vc;
    k21   = Q/Vp;
    Cc    = centr/Vc;
    d/dt(centr)  = - ke*centr - k12*centr + k21*periph;
    d/dt(periph) =            + k12*centr - k21*periph;
    d/dt(AUC)    =   Cc;
  })
```

## error_model
A function of the residual error model, taking two arguments: the simulated concentrations, and a vector `sigma` of the estimates of the parameters for the residual error model.

```{r error_model}
error_model <- function(f,sigma){           #additive model if sigma[2] == 0
  g <- sigma[1]^2 + (sigma[2]^2)*(f^2)      #proportional model if sigma[1] == 0
  return(sqrt(g))
}
```

For multiple endpoints models, `error_model` must be a named list with a function for each endpoint `vignette("multiple_endpoints")`.

```{r error_model_multiple_endpoints}
error_model = list(
    first_endpoint = function(f,sigma){
      g <- sigma[1]^2 + (sigma[2]^2)*(f^2)
      return(sqrt(g))
    },
    second_endpoint = function(f,sigma){
      g <- sigma[1]^2 + (sigma[2]^2)*(f^2)
      return(sqrt(g))
    }
  )
```

To obtain individual estimations for multiple endpoints, consistency in the naming convention must be maintained across the following:

  *  The dataset (using the column DVID).
  *  The residual error models (stored in a named list).
  *  The standard deviation of the residual error models (stored in a named list, see **sigma**).

As many residual error models as desired can be defined. Each model defined in the named list error_models must have its counterpart in the named list `sigma`, and the names must match those defined in the `DVID` column of the dataset.

## theta
The estimations of the parameters for the fixed effects of the model (THETA), in a named vector. The names must match the names used in `ppk_model`.

```{r theta}
theta = c(THETA_Cl=4.5, THETA_Vc=58.4, THETA_Vp=38.4, THETA_Q=6.5)
```

## omega
The variance-covariance matrix of the random effects (ETA) for the individual model. A symmetric matrix. The names must match the names used in `ppk_model`. An easy way to define it is using `lotri::lotri()`.

The estimates of the variances of the random effects can be given under different parameterizations depending on the authors.

* Standard deviation (SD): the square root of the variance, as returned by Monolix
* Coefficient of variation (CV): calculated as `sqrt(exp(SD^2)-1)`, the variance can be computed back with `log((CV^2)+1)`
* Full covariance matrix: the easiest to reuse, but less common in the literature

In the case of the [vancomycin model](https://doi.org/10.1097/FTD.0000000000000490), the estimates of between subject variability (BSV) are given as CV%. They must be converted to variances prior to their inclusion in `omega`.

|Parameter |CV%  (from the article)|Variance = SD^2|
|----------|:---------------------:|:-------------:|
|BSV on CL |39.8                   | 0.147         |
|BSV on Vc |81.6                   | 0.510         |
|BSV on Vp |57.1                   | 0.282         |

```{r omega}
omega = lotri::lotri({ETA_Cl + ETA_Vc + ETA_Vp + ETA_Q ~
                          c(0.147,
                            0    ,  0.510 ,
                            0    ,  0     ,   0.282,
                            0    ,  0     ,   0    ,    0)})

```

The estimates of covariance (off-diagonal) are sometimes given as coefficients of correlation between ETAs. The covariance between ETA_a and ETA_b can be computed with the following product: `standard_deviation(ETA_a) * standard_deviation(ETA_b) * correlation(ETA_a and ETA_b)`.

In this example, all covariances are equal to zero.

## sigma
The estimates of the parameters for the residual error model on the standard deviation scale, either in a vector:

```{r sigma}
sigma       = c(additive_a = 3.4, proportional_b = 0.227)
```

in a matrix:

```{r matrix}
sigma       = lotri::lotri({prop + add ~ c(0.227,0.0,3.4)})
```

or in a named list (for multiple endpoints):

```{r named_list}
sigma       = list(
    first_endpoint=c(additive_a = 0.144, proportional_b = 0.15),
    second_endpoint=c(additive_a = 3.91, proportional_b = 0.0)
    )
```

depending on the residual error model.

## pi_matrix
Optional: only needed for models with inter-occasion variability (IOV).
The variance-covariance matrix of the random effects (KAPPA) for the IOV. As for the `omega` matrix, the names must match the names used in `ppk_model`. An easy way 
to define it is using `lotri::lotri()`.

```{r pi_matrix}
pi_matrix = lotri::lotri({KAPPA_Cl + KAPPA_Vc ~
      c(0.1934626,
        0.00     ,  0.05783106)})

```

## covariates
The names of every covariate defined in `ppk_model`, in a character vector.

```{r covariates}
covariates  = c("CLCREAT","WT","DIAL")
```

## Full model
The `posologyr` model is the list of all these objects. 
Note: this model does not include inter-occasion variability, so the pi_matrix is omitted.

```{r vancomycin_pososlogyr_model_list}
mod_vancomyin_Goti2018 <- list(
  ppk_model   = rxode2::rxode({
     centr(0) = 0;
    TVCl  = THETA_Cl*(CLCREAT/120)^0.8*(0.7^DIAL);
    TVVc  = THETA_Vc*(WT/70)          *(0.5^DIAL);
    TVVp  = THETA_Vp;
    TVQ   = THETA_Q;
    Cl    = TVCl*exp(ETA_Cl);
    Vc    = TVVc*exp(ETA_Vc);
    Vp    = TVVp*exp(ETA_Vp);
    Q     = TVQ;
    ke    = Cl/Vc;
    k12   = Q/Vc;
    k21   = Q/Vp;
    Cc    = centr/Vc;
    d/dt(centr)  = - ke*centr - k12*centr + k21*periph;
    d/dt(periph) =            + k12*centr - k21*periph;
    d/dt(AUC)    =   Cc;
  }),
  error_model = function(f,sigma){
    g <- sigma[1] + sigma[2]*f
    return(g)
  },
  theta = c(THETA_Cl=4.5, THETA_Vc=58.4, THETA_Vp=38.4,THETA_Q=6.5),
  omega = lotri::lotri({ETA_Cl + ETA_Vc + ETA_Vp + ETA_Q ~
                          c(0.147,
                            0    ,  0.510 ,
                            0    ,  0     ,   0.282,
                            0    ,  0     ,   0    ,    0)}),
  sigma       = c(additive_a = 3.4, proportional_b = 0.227),
  covariates  = c("CLCREAT","WT","DIAL"))
```

### Resulting R object

```{r vancomycin_pososlogyr_model}
mod_vancomyin_Goti2018
```
