---
title: "Route of administration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Route of administration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
library(rxode2)
setRxThreads(2L)  # limit the number of threads
```

```{r setup}
library(posologyr)
```

# Introduction
The Caldès 2009 ganciclovir model (https://doi.org/10.1128/aac.00085-09) is capable of describing the pharmacokinetics of either injectable ganciclovir or oral valganciclovir.

```{r model}
mod_ganciclovir_Caldes_2009 <- function() {
  ini({
    THETA_cl  <- 7.49
    THETA_v1  <- 31.90
    THETA_cld <- 10.20
    THETA_v2  <- 32.0
    THETA_ka  <- 0.895
    THETA_baf <- 0.825
    ETA_cl ~ 0.107
    ETA_v1 ~ 0.227
    ETA_ka ~ 0.464
    ETA_baf ~ 0.049
    add.sd <- 0.465
    prop.sd <- 0.143
  })
  model({
    TVcl  = THETA_cl*(ClCr/57);
    TVv1  = THETA_v1;
    TVcld = THETA_cld;
    TVv2  = THETA_v2;
    TVka  = THETA_ka;
    TVbaf = THETA_baf;

    cl  = TVcl*exp(ETA_cl);
    v1  = TVv1*exp(ETA_v1);
    cld = TVcld;
    v2  = TVv2;
    ka  = TVka*exp(ETA_ka);
    baf = TVbaf*exp(ETA_baf);

    k10 = cl/v1;
    k12 = cld / v1;
    k21 = cld / v2;
    Cc = centr/v1;

    d/dt(depot)  = -ka*depot
    d/dt(centr)  =  ka*depot - k10*centr - k12*centr + k21*periph;
    d/dt(periph) =                         k12*centr - k21*periph;
    d/dt(AUC)    = Cc;

    f(depot)=baf;
    alag(depot)=0.382;

    Cc ~ add(add.sd) + prop(prop.sd) + combined1()
  })
}

mod_ganciclovir_Caldes_2009 <- mod_ganciclovir_Caldes_2009()
```

# Intravenous ganciclovir

## Patient record with TDM data

To describe intravenous administration, a CMT column has been added to the TDM data table to indicate administrations directly into the central compartment.

Note: to compute the AUC between the last dose and the time of the last dose + 24 hours, a dummy dose of 0 mg is added to the time of the last observation of interest (i.e. H144).

```{r tdm_patientA}
patient <- data.frame(ID=1,TIME=c(0,121,122,126,144),
                      DV=c(NA,10.8,5.8,3.3,NA),
                      ADDL=c(5,0,0,0,0),
                      II=c(24,0,0,0,0),
                      EVID=c(1,0,0,0,1),
                      CMT=c("centr",NA,NA,NA,"centr"),
                      AMT=c(250,0,0,0,0),
                      DUR=c(0.5,NA,NA,NA,NA),
                      ClCr=25)
patient
```

## Individual PK profile and AUC 0-24

The individual PK profile can be estimated, and plotted.

```{r estim_map}
map_patient <- poso_estim_map(patient,mod_ganciclovir_Caldes_2009)
```

```{r plot_map}
#| fig.alt: >
#|   Plot of the individual PK profile
plot(map_patient$model,Cc)
```

The difference between the cumulative AUC at H144 and that at H120 gives the AUC 0-24 after the last dose. Using data.table is optional, but the syntax is more convenient.

```{r auc_24_ganci}
library(data.table)
data.table(map_patient$model)[time==144,AUC] - 
  data.table(map_patient$model)[time==120,AUC]
```

## Optimal dose for an intravenous ganciclovir injection

The optimal dose to achieve an AUC of 50 mg.h/L can be determined for a new injection of IV ganciclovir by setting `cmt_dose = "centr"`.

```{r optimal_intravenous_ganci}
poso_dose_auc(patient,mod_ganciclovir_Caldes_2009,tdm=TRUE,
              time_dose = 145,
              duration = 1,
              time_auc = 24,
              target_auc = 50,
              cmt_dose = "centr")
```

## Optimal dose for an oral valganciclovir administration

The optimal dose to achieve an AUC of 50 mg.h/L can be determined for an administration of oral valganciclovir by setting `cmt_dose = "depot"`. 

```{r optimal_oral_valganci}
poso_dose_auc(patient,mod_ganciclovir_Caldes_2009,tdm=TRUE,
              time_dose = 145,
              time_auc = 24,
              target_auc = 50,
              cmt_dose = "depot")
```
Keeping the default value of cmt_dose, which is the first compartment declared in the PK model, would also work here.

> **Note** in this case the estimated dose should be corrected by a factor of 1/0.72 as the Caldès 2009 paper states: 
>
>
>*"As NONMEM estimated the pharmacokinetic parameters of ganciclovir, valganciclovir doses were converted to their equivalent ganciclovir content by multiplying the valganciclovir dose by 0.720 (corresponding to the ratio between the molecular weights of ganciclovir and valganciclovir)."*
